\documentclass[12pt]{article}

\usepackage{amssymb,amsmath,amsthm}

% theorem styles (I like everything to have the same counter)
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{exercise}[theorem]{Exercise}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% some numbering settings
\numberwithin{equation}{section}

\usepackage{fancyhdr}
\usepackage{lastpage}
\setlength{\headheight}{15.2pt}
\renewcommand{\footrulewidth}{0.4pt}% default is 0pt
\setlength{\footskip}{30pt}
\pagestyle{fancy}

\makeatletter
\let\ps@plain\ps@fancy 
\makeatother

\lhead{MATH 742}
\chead{Representations over non-closed fields}
\rhead{Spring 2023}
\lfoot{Last Revised: \today}
\cfoot{}
\rfoot{\thepage\ of \pageref{LastPage} }

\begin{document}

%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%
\title{Representations over non-closed fields}
\author{Alexander Duncan}

\maketitle

Throughout, $k$ is a field of characteristic $0$
and $\overline{k}$ is the algebraic closure of $k$.
(For those uncomfortable with infinite Galois extensions,
one can instead replace $\overline{k}$ with the Galois closure
of the finitely many relevant finite field extensions
showing up in any of the contexts we discuss below.)

Much of this is inspired from \cite[\S{12--13}]{Serre}.

\section{Representation Ring via Modules}

Let $G$ be a finite group.

Recall that $kG$ is the group algebra of $G$ over $k$.
Since $k$ has characterstic $0$, we know $kG$ is semisimple.
We have a canonical decomposition
\[
kG \cong \bigoplus_{i=1}^r A_i
\]
where $A_1,\ldots,A_r$ are simple $k$-algebras.

Each $A_i$ is a central simple $F_i$-algebra
for a finite field extension $F_i/k$.
Let $n_i = \deg(A_i)$ be the \emph{degree} of $A_i$.
Equivalently, we have $n_i^2 = \dim_{F_i}(A_i)$
or
\[
A_i \otimes_{F_i} \overline{k} \cong
\operatorname{M}_{n_i}(\overline{k})
\]
as $\overline{k}$-algebras.

More precisely, there exists a central division $F_i$-algebra $D_i$
and a positive integer $m_i$ such that
\[
A_i \cong \operatorname{M}_{m_i}(D_i)
\]
as $F_i$-algebras.
Both $m_i$ and the isomorphism class of $D_i$ are uniquely determined.
Let $d_i := \operatorname{ind}(A_i)$ denote the Schur index of $A_i$.
We have $d_i = \deg(D_i)$, which is equivalent to
$d_i^2=\dim_{F_i}(D_i)$.  Also, we have $n_i=m_id_i$.

Each $A_i$ has a simple module $V_i$, which is unique up to isomorphism.
Since $V_i \cong D_i^{\oplus n_i}$, we see that
\[
\dim_k V_i = [F_i:k]m_id_i^2 = [F_i:k]d_in_i .
\]
The vector spaces $V_1,\ldots,V_r$ are precisely the
irreducible representations of $G$ over $k$.
The \emph{Schur index} of $V_i$ is defined to be the
Schur index $d_i$ of the associated algebra $A_i$.

Let $\Gamma$ be the absolute Galois group
$\operatorname{Gal}(\overline{k}/k$.
Let $X_i :=
\operatorname{Hom}_{k-\operatorname{alg}}(F_i,\overline{k})$.
Observe that $X_i$ has a left action of $\Gamma$ by
post-composition.
Informally, $X_i$ can be thought of as the set of roots of the minimal
polynomial of a primitive element of $F_i/k$.
We have a $\overline{k}$-algebra isomorphism
\[
\omega : F_i \otimes_k \overline{k} \cong
\bigoplus_{\sigma \in X_i} \overline{k}
\]
via $\omega(f \otimes \ell)_{\sigma} := \sigma(f)\ell$.

For each $\sigma \in X_i$, we may define a tensor product
$A_i \otimes_\sigma \overline{k}$ of $F_i$-algebras
where we use the structure map $F_i \to A_i$ on the left and
the homomorphism $\sigma : F_i \to \overline{k}$ on the right.
Via the sequence of isomorphisms of $\overline{k}$-algebras
\[
A_i \otimes_k \overline{k} \cong 
A_i \otimes_{F_i} F_i \otimes_k \overline{k}
\cong A_i \otimes_{F_i} \bigoplus_{\sigma \in X_i} \overline{k}
\]
we obtain an isomorphism
\[
\Omega : A_i \otimes_k \overline{k} \cong
\bigoplus_{\sigma \in X_i} A_i \otimes_\sigma \overline{k} .
\]

Let $W_{i,\sigma}$ be the irreducible representation
of $\overline{k}G$ corresponding to $A_i \otimes_\sigma \overline{k}$.
We conclude that we have the following decomposition
\[
V_i \otimes_k \overline{k} \cong
\bigoplus_{\sigma \in X_i} W_{i,\sigma}^{\oplus d_i}
\]
for each irreducible representation $V_i$ of $kG$.

If $\rho_{i,\sigma} : G \to \operatorname{GL}(W_{i,\sigma})$ are
the corresponding maps, then notice that
\[
\tau \circ \rho_{i,\sigma} = \rho_{i,\tau(\sigma)}
\]
for every $\tau \in \Gamma$.
Thus, there is a left action of $\Gamma$ on the set of irreducible
representations of $G$ over $\overline{k}$
with orbits corresponding to the $\Gamma$-sets $X_i$.

\section{Character Theory}

Here we will assume that $\overline{k}$ (and therefore $k$)
is a specific subfield of $\mathbb{C}$.
This is essentially harmless in view of the Lefschetz principle.
For the suspicious, assume that $\mathbb{C}$ is just bizarre notation for
$\overline{k}$ throughout the remainder of the document.

Recall that the representation ring $R_k(G)$ of $G$ over $k$
is the additive group of virtual representations over $k$
with multiplication obtained extending the tensor product
on the subrig $R_k^+(G)$.
From above, we see that $R_k(G)$ is a free abelian group
on the set $[V_1],\ldots,[V_r]$.

The map $[V] \mapsto [V \otimes_k \overline{k}]$
gives a canonical injective ring homomorphism
\[
R_k(G) \hookrightarrow R_{\mathbb{C}} = R(G)
\]
and therefore a canonical injective ring homomorphism
\[
R_k(G) \hookrightarrow C(G)
\]
where $C(G)$ is the set of complex class functions on $G$.

Just as in the complex case, given a representation $(V,\rho)$
of $G$ over $k$, we define the character
$\chi_V : G \to k$
via the trace $\chi_V(g) := \operatorname{tr}(\rho(g))$.
This agrees with the map $R_k(G) \hookrightarrow C(G)$ defined above.
Thus, we identify $R_k(G)$ with a subring of $R(G)$ and $C(G)$ in what
follows.

Let $\chi_1,\ldots,\chi_r$ be the irreducible representations of $G$
over $k$ corresponding to the algebras $A_1,\ldots, A_r$ above.
Let $\psi_{i,\sigma} : G \to \overline{k}=\mathbb{C}$ be the character
corresponding to the irreducible representation $W_{i,\sigma}$
defined in the previous section.

\begin{theorem}
For every $1 \le i \le r$, we have
\[
\chi_i = d_i \sum_{\sigma \in X_i} \psi_{i,\sigma} .
\]
In particular, the characters $\chi_1,\ldots, \chi_r$
are an orthogonal basis for the subspace $R_k(G)$ of $C(G)$.
\end{theorem}

\begin{proof}
The description of $\chi_i$ follows from the results in the previous
section.
The orthogonality follows from the fact that the $\psi_{i,\sigma}$
are orthogonal.
\end{proof}

Note that we only have the $\{\chi_1,\ldots,\chi_r\}$ is an
\emph{orthogonal} basis; it is not necessarily orthonormal.

There are in fact \emph{three} different traces that one may consider
for each irreducible representation $V_i$.
We have the ordinary trace $\chi_i(g) \in k$ as an element of
$\operatorname{End}_k(V_i)$.
We may also define $\phi_i(g) \in F_i$ as the trace of $g$ in
$\operatorname{End}_{F_i}(V_i)$.
Finally, we have $\psi_i(g) \in F_i$ as the reduced trace
$\operatorname{Trd}(g)$ viewing $g$ as an element of the central simple
$F_i$-algebra $A_i$.
These traces are all related via
\[
\chi_i = \operatorname{Tr}_{F_i/k} \circ \phi_i, \quad
\phi_i = d_i \psi_i
\]
where we recall that $d_i$ is the Schur index of the central simple
$F_i$-algebra $A_i$ (equivalently, the representation $V_i$).
Finally for each $\sigma \in X_i$, we see that $\psi_{i,\sigma} = \sigma
\circ \psi_i$.
This gives another way of obtaining the decomposition of $V_i \otimes_k
\overline{k}$ above.

\subsection{Definability of Representations}

Recall, from the previous section, that the absolute Galois group
permutes the irreducible representations of $\overline{k}G$.
Indeed, if $\rho_{i,\sigma} : G \to \operatorname{GL}(W_{i,\sigma})$
is the map then we have
\[
\tau \circ \rho_{i,\sigma} = \rho_{i,\tau(\sigma)}
\]
for $\tau \in \Gamma$.
This action is identical to the one obtained by acting on the
characters
\[
\tau \circ \phi_{i,\sigma} = \phi_{i,\tau(\sigma)}
\]
where we recall $\phi_{i,\sigma}: G \to \overline{k}$ is just a
particular class function.

\begin{theorem}
Let $\overline{R}_k(G)$ be the subset of $R(G)$ consisting of characters
whose class functions whose image is defined over $k$.
Then $\overline{R}_k(G)$ has basis
\[
\frac{\chi_1}{d_1},
\frac{\chi_2}{d_2},
\cdots,
\frac{\chi_r}{d_r}.
\]
\end{theorem}

Thus $\overline{R}_k(G)$ can be completely recovered from the character
table over $\mathbb{C}$.  Namely, one can reconstruct the $\chi_i/d_i$
as the sums of $\Gamma$-orbits of the complex characters.
To reconstruct $\overline{R}_k(G)$, one needs to know the Schur indices.

\begin{proposition}
If $G$ is an abelian group,
then $R_k(G) = \overline{R}_k(G)$
\end{proposition}

\begin{proof}
If $G$ is abelian, then $kG$ is commutative, so all constituent
division algebras are fields.  Thus all Schur indices are trivial.
\end{proof}

We are now in a position to prove an important theorem of Brauer, which
essentially reduces the study of definability of representations of
arbitrary fields of characteristic $0$ to the case of cyclotomic fields.

\begin{theorem} \label{thm:define_over_cyclotomic}
If $G$ is a finite group of exponent $e$,
then every representation of $G$ over an arbitrary field of
characteristic $0$
is conjugate to a representation defined over the
cyclotomic field $\mathbb{Q}(\zeta_e)$.
\end{theorem}

\begin{proof}
Let $K=\mathbb{Q}(\zeta_e)$.  We want to show that $R_K(G)=R(G)$.
By Brauer's theorem, we may write a general element $\chi$ of $R(G)$
in the form
\[
\rho = \sum_{i=1}^m c_i\operatorname{Ind}_{H_i}^G(\tau_i)
\]
where the $c_i$'s are integers, the $H_i$ are subgroups of $G$,
and each $\tau_i$ is a one-dimensional representations of $R(H_i)$.
A one-dimensional representation of $H_i$ is defined over
$\mathbb{Q}(\zeta_e)$ since $H_i$ has exponent at most $e$.
We conclude that $\chi$ is an integral linear combination of
representations induced from $R_K(H_i)$, so is in $R_K(G)$ as desired.
\end{proof}

\section{Column Orthogonality}

We saw above that $\chi_1,\ldots,\chi_r$ are an orthogonal basis for
$R_k(G)$.  This amounts to showing that the \emph{rows} of the character
table of $G$ (over $k$) are orthogonal.
However, we now have ``too many'' columns if we expect a square
character table.
Here we investigate this issue.

For each positive integer $n$, define the \emph{Adams operation}
$\Psi^n : C(G) \to C(G)$ as the function that takes
a class function $\chi$ to the class function $\Psi^n(\chi)$ defined by
\[
\Psi^n(\chi)(g) := \chi(g^n)
\]
for each $g \in G$.

\begin{proposition}
Every Adams operation restricts to a ring homomorphism
from $R_k(G)$ to itself.
\end{proposition}

\begin{proof}
Let $\chi$ be the character of a representation $(V,\rho)$.
For a given $g$, consider the multiset of eigenvalues
\[
a_1,\ldots, a_m
\]
of $\rho(g)$.
Thus $\chi(g)=p_1$ where $p_1$ is the first power sum polynomial
in $a_1,\ldots, a_m$.
We have $\chi(g^n)=p_n$ where $p_n$ is the power sum polynomial
\[
a_1^n + \cdots + a_m^n.
\]
By the fundamental theorem of symmetric polynomials, there exists a
polynomial $f_n$ with integer coefficients such that
\[
p_n = f_n(e_1,\ldots,e_m)
\]
where $e_1,\ldots,e_m$ are the elementary symmetric polynomials
in $a_1,\ldots,a_n$.

The function $f_n$ does not depend on the eigenvalues, so we have an
expression
\[
\Psi^n(\chi) = f_n(\Lambda^1 \chi, \ldots, \Lambda^n \chi)
\]
in the ring $C(G)$. Since exterior powers of a representation are
representations, we see that $\Psi^n(\chi)$ is in $R_k(G)$ as desired.
\end{proof}

Recall that the \emph{exponent} $e$ of a finite group is the least common multiple
of the orders of all the elements $g \in G$.
Equivalently, the exponent $e$ is the least positive integer such that
$g^e=1$ for all $g \in G$.

\begin{lemma}
If $n$ is relatively prime to $e$,
then the function $g \mapsto g^t$ is a bijection from $G$ to itself,
which takes conjugacy classes to conjugacy classes.
\end{lemma}

\begin{proof}
Let $s$ be the multiplicative inverse of $n$ modulo $e$,
which exists since $t$ is relatively prime to $e$.
The function $g \mapsto g^s$ is the inverse function to
$g \mapsto g^t$ since $g^{ts}=g^1=g$.
\end{proof}

From this we immediately conclude:

\begin{corollary}
If $t$ is relatively prime to $e$,
then $\Psi^t : R(G) \to R(G)$ is an isomorphism.
\end{corollary}

Thus, for appropriate $t$, the Adams operations $\Psi^t$
permute the \emph{columns} of the character table of $R(G)$.

In view of Theorem~\ref{thm:define_over_cyclotomic},
we may assume that $k \subseteq K=\mathbb{Q}(\zeta_e)$
where $e$ is the exponent of $G$.
Let $\Gamma$ be the Galois group of $K/k$.
Observe that $\Gamma$ is a subgroup of
$(\mathbb{Z}/e\mathbb{Z})^\times$.
Let $\Gamma$ be generated by $\sigma_1,\ldots,\sigma_s$
where $\sigma(\zeta_e)=\zeta_e^t$ for some $t_1,\ldots,t_s$ in
$(\mathbb{Z}/e\mathbb{Z})^\times$.

We now have two actions of $\Gamma$ on $R(G)$.
Given $\gamma \in \Gamma$ and $\chi \in R(G)$, we have
\[
(\gamma \ast \chi)(g) := \gamma(\chi(g)) 
\]
for all $g \in G$,
where we view $\chi: G \to K$ as a function with codomain $K \subset
\overline{k}$.
We also have the action using the Adams operation
which is generated by
\[
\sigma_i \ast' \chi := \Psi^{t_i}(\chi)
\]
for $1 \le i \le s$.

\begin{proposition}
The two actions of $\Gamma$ on $R(G)$ coincide.
\end{proposition} 

\begin{proof}
The eigenvalues of every matrix $M$ in every representation of $G$ is an
$e$th root of unity.
Thus, if $\chi$ is a character in $R(G)$, then
\[
(\sigma_i \circ \chi)(g) = \Psi^{t_i}(\chi)(g)
\] 
for every $\sigma \in \Gamma$.
\end{proof}

Observe that we can define the $\Gamma$-action directly on the conjugacy
classes of $G$ using the above identification.
In view of this, we define the \emph{$\Gamma$-classes}
(or \emph{$K$-classes}) of $G$ to be the unions of $\Gamma$-orbits of
conjugacy classes.
We therefore have:

\begin{theorem}
$\overline{R}_k(G)$ is precisely the subset of $R(G)$ whose corresponding
characters are fixed by the $\Gamma$-action.
\end{theorem}

As an immediate corollary, we obtain:

\begin{corollary}
The number of irreducible characters of $G$ over $k$
is equal to the number of $\Gamma$-classes of $G$.
\end{corollary}

Note that the particular action of $\Gamma$ on $G$ is determined by a
choice of isomorphism of $\Gamma$ with its image in
$(\mathbb{Z}/e\mathbb{Z})^\times$, but the $\Gamma$-classes do not
depend on this choice.
Thus the $\Gamma$-classes of $G$ are totally group-theoretic.
In particular, we don't actually need to compute the character table
to determine the number of irreducible representations (just like over
$\mathbb{C}$).

However, we only obtain $\overline{R}_k(G)$ from this analysis.
More work needs to be done to identify the Schur indices.

\subsection{Representations over the rationals}

Here we consider the important special case where $k=\mathbb{Q}$.
Let $G$ be a finite group with exponent $e$.
As above, $K=\mathbb{Q}(\zeta_e)$ where $\zeta_e$ is a primitive $e$th
root of unity.
The Galois group $\Gamma = \operatorname{Gal}(K/k)$ is isomorphic
to the full group $(\mathbb{Z}/e\mathbb{Z})^\times$ by standard results
on cyclotomic fields.

In this case, the $\Gamma$-classes from above have a simple description:

\begin{lemma}
Two elements $g,h \in G$ belong to the same $\mathbb{Q}$-class if and
only if they generate conjugate cyclic subgroups.
\end{lemma}

\begin{proof}
Observe that $\langle g \rangle = \langle h \rangle$ if and only if
$g^s = h$ and $h^t= g$ for some integers $s$ and $t$.
Observe that $t,s$ are relatively prime to $n=\langle g \rangle$.
Since $n$ divides the exponent $e$ of $G$, we have a surjective group
homomorphism
$(\mathbb{Z}/e\mathbb{Z})^\times \to (\mathbb{Z}/n\mathbb{Z})^\times$.
Thus, we may assume $s$ and $t$ are relatively prime to $e$.
Since $\Gamma \cong (\mathbb{Z}/e\mathbb{Z})^\times$,
we conclude $\langle g \rangle = \langle h \rangle$ if and only if
$g$ and $h$ are in the same $\Gamma$-orbit.
This respects conjugacy classes. 
\end{proof}

We conclude the following:

\begin{theorem}
The number of isomorphism classes of irreducible representations of a
finite group $G$ over $\mathbb{Q}$ is equal to the number of conjugacy
classes of cyclic subgroups of $G$.
\end{theorem}

We point out the following pleasant parallels with permutation
representations and complex representations:

\begin{itemize}
\item The number of irreducible \textbf{permutation} representations are in
bijection with the number of conjugacy classes of \textbf{subgroups}.
\item The number of irreducible \textbf{rational} representations are in
bijection with the number of conjugacy classes of \textbf{cyclic
subgroups}.
\item The number of irreducible \textbf{complex} representations are in
bijection with the number of conjugacy classes of \textbf{elements}.
\end{itemize}

\section{Representations over the reals}

Let $V$ be a finite-dimensional
complex representation of a finite group $G$.

\begin{proposition}
$V$ has a $G$-invariant inner product.
\end{proposition}

\begin{proof}
Let $b : V \times V \to \mathbb{C}$ be an inner product (not necessarily
$G$-invariant).
Define a function $B : V \times V \to \mathbb{C}$ by
\[
B(v,w) = \sum_{g \in G}b(gv,gw)
\]
for $v, w \in V$.
Observe that $B$ is also sesquilinear
and that $B(v,v) > 0$ for all $v \ne 0$.
Thus $B$ is also an inner product.
\end{proof}

The previous proposition has the following interpretation:

\begin{corollary}
Every finite subgroup of $\operatorname{GL}_n(\mathbb{C})$
is conjugate to a finite subgroup of the unitary group $\operatorname{U}_n$.
\end{corollary}

Crucially, we can only produce a \emph{sesquilinear} form in general;
there may not be a non-trivial \emph{bilinear} form.
For example, a one-dimensional complex representation of
$\mathbb{Z}/3\mathbb{Z}$ has no non-trivial bilinear forms.
However, we know that sesquilinear forms and bilinear forms are
essentially the same over real closed fields.
It turns out this leads to an interesting theory. 

We say that $V$ is \emph{realizable over $\mathbb{R}$} if there exists a
real representation $W$ of $G$ such that $W \otimes_\mathbb{R}
\mathbb{C} \cong V$.

\begin{theorem}[Frobenius-Schur] \label{thm:FrobSchur}
A complex representation $V$ is realizable over $\mathbb{R}$ if and only
if there exists a $G$-invariant non-degenerate symmetric bilinear form
on $V$.
\end{theorem}

\begin{proof}
(The proof is short, but is omitted;
see \cite[Theorem 31]{Serre}.)
\end{proof}

Now assume $V$ is irreducible and let $\chi$ be its character.

\begin{definition}
The \emph{Frobenius-Schur indicator} of $V$ is the number
\[
\iota(V) := \frac{1}{|G|}\sum_{g \in G} \chi(g^2).
\]
\end{definition}

In order to understand representation theory over $\mathbb{R}$, it is
worth pointing out that there are only three division
$\mathbb{R}$-algebras:
\begin{itemize}
\item the real field $\mathbb{R}$,
\item the complex field $\mathbb{C}$, and
\item Hamilton's quaternions $\mathbb{H}$.
\end{itemize}
The irreducible representation $V$ is a simple module of
$A \otimes_{\mathbb{R}} \mathbb{C}$ where $A$ is a simple subalgebra of
$\mathbb{R}G$ and $A \cong \operatorname{M}_n(D)$ where $D$ is one of
the three cases above.

Thus, we get a trichotomy of the real behavior of complex
representations.  Let us list some of their features.

\noindent
\textbf{Complex:}
\begin{itemize}
\item the division algebra corresponding to $V$ is $\mathbb{C}$,
\item Frobenius-Schur indicator is $\iota(V)=0$,
\item $V$ does not have a $G$-invariant non-degenerate bilinear form,
\item $\chi$ is not real valued,
\item $V$ is not realizable over $\mathbb{R}$, and
\item $V$ is not isomorphic to its dual $V^\vee$.
\end{itemize}

\noindent
\textbf{Real:}
\begin{itemize}
\item the division algebra corresponding to $V$ is $\mathbb{R}$,
\item Frobenius-Schur indicator is $\iota(V)=1$,
\item $V$ has a $G$-invariant non-degenerate symmetric bilinear form,
\item $\chi$ is real valued,
\item $V$ is realizable over $\mathbb{R}$, and
\item $V \cong V^\vee$.
\end{itemize}

\noindent
\textbf{Quaternionic:}
\begin{itemize}
\item the division algebra corresponding to $V$ is $\mathbb{H}$,
\item Frobenius-Schur indicator is $\iota(V)=-1$,
\item $V$ has a $G$-invariant non-degenerate skew-symmetric bilinear form,
\item $\chi$ is real valued,
\item $V$ is not realizable over $\mathbb{R}$, and
\item $V \oplus V$ is realizable over $\mathbb{R}$, and
\item $V \cong V^\vee$.
\end{itemize}

We merely sketch why the above characterization holds,
leaving details to the reader (or \cite[\S{13.2}]{Serre}).
Most of the characterizations follow quickly from Wedderburn theory,
so we focus on the bilinear forms.

Note that bilinear forms are equivalent to maps between $V$ and
$V^\vee$, where non-triviality is equivalent to non-degeneracy by
Schur's Lemma since we want $G$-invariance.
Schur's Lemma also tells us $G$-invariant bilinear forms
are unique up to scaling.
The decomposition of a bilinear form into a symmetric and skew-symmetric
part is $G$-invariant, so uniqueness up to scaling implies that the form
must be either symmetric or skew-symmetric.
Having a \emph{symmetric} bilinear form is equivalent to being realizable over
the reals by Theorem~\ref{thm:FrobSchur} above.

Finally, we consider the Frobenius-Schur indicator.
A $G$-invariant symmetric bilinear form on $V$ is equivalent to
having a trivial subrepresentation of $\operatorname{Sym}^2(V)$.
Similarly, a $G$-invariant skew-symmetric bilinear form on $V$
is equivalent to having a trivial subrepresentation of
$\operatorname{Alt}^2(V)$.
Recall the formulas for the characters of the
symmetric and alternating squares:
\[
\chi_{\operatorname{Sym}^2(V)} = \chi_V^2 + \Psi^2(\chi_V)
\textrm{ and }
\chi_{\operatorname{Alt}^2(V)} = \chi_V^2 - \Psi^2(\chi_V).
\]
Thus, we have $(1,\chi_{\operatorname{Sym}^2(V)})=1$ iff $V$ is real,
$(1,\chi_{\operatorname{Alt}^2(V)})=1$ iff $V$ is quaternionic,
and they are $0$ otherwise.

Now we simply observe that
\[
\chi_V^2 = \chi_{\operatorname{Sym}^2(V)} + \chi_{\operatorname{Alt}^2(V)}
\]
and
\[
\iota(V) := \frac{1}{|G|}\sum_{g \in G} \chi(g^2)
=(1, \Psi^2(\chi_V)).
\]

\bibliographystyle{alpha}
\bibliography{rep_theory}

\end{document}


